Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
Change & harmonise notation for branch length

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

This paper presents one way of generalizing of this idea.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

\begin{definition}[Sample and subtree weights] \label{defn:weights}
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these and a tree $T$,
    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
    of every sample node that is descended $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
    The \emph{total weight} is the sum of the weights over all samples:
    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
    that contains all samples.
\end{definition}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry --
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.

>>>>

\jk{These two intro paragraphs aren't working for me. There's too many
parenthetical comments, mixing high-level overview with getting into
the weeds. Does the first para actually help? Perhaps we're just confusing
people by talking about the AFS here?}
Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

% 2-3 sentances of intro before this?
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

\begin{definition}[Sample and subtree weights] \label{defn:weights}
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these and a tree $T$,
    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
    The \emph{total weight} is the sum of the weights over all samples:
    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
    that contains all samples.
\end{definition}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $\beta_T(u)$.
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry ---
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.
